---
title: "Weighted networks: strength or distance?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted networks: strength or distance?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most real networks come with a number on every edge: a confidence score from an
interaction database, a correlation from a co-expression matrix, a count of
supporting publications, a reaction rate. This article is about what netkit does
with that number, and why it asks you to say what kind of number it is.

```{r setup}
library(netkit)
suppressMessages(library(igraph))
```

```{r helper, include = FALSE}
# netkit resolves `weights` and `weight_type` into a strength vector and a cost
# vector internally. This vignette peeks at that conversion to show what the two
# readings actually do; it is not part of the public interface.
as_netkit_weights_demo <- function(graph, attr, type) {
  w <- edge_attr(graph, attr)
  if (type == "strength") {
    list(strength = w, distance = ifelse(w == 0, Inf, 1 / w))
  } else {
    list(strength = max(w) - w + min(w[w > 0]), distance = w)
  }
}
```

## The problem

A weight can mean one of two opposite things.

A **strength** means *larger values bind the endpoints more tightly*. A
BioGRID confidence score of 0.9 is a stronger claim than 0.2. A co-expression
correlation of 0.95 is a closer relationship than 0.3.

A **distance** means *larger values push the endpoints further apart*. A
reaction time, a transit cost, a dissimilarity.

Which one you have changes the answer to almost every network question, in
opposite directions. And `igraph` — which netkit is built on — reads the `weight`
edge attribute implicitly, giving it **different meanings in different
functions**: a cost in `betweenness()`, `distances()`, `diameter()` and
`mean_distance()`, but a strength in `cluster_louvain()` and the other community
detection algorithms.

Here is what that looks like. The graph is a barbell: two cliques joined by a
single bridge.

```{r barbell}
g <- disjoint_union(make_full_graph(4), make_full_graph(4))
V(g)$name <- c(paste0("L", 1:4), paste0("R", 1:4))
g <- add_edges(g, c("L1", "R1"))

# A confidence score: the bridge is a very well-supported interaction, the
# within-clique edges less so.
E(g)$confidence <- 0.1
E(g)$confidence[E(g)["L1" %--% "R1"]] <- 1.0
```

Read as a strength, that bridge is the most reliable edge in the graph, so
crossing it should be *cheap*. Read as a cost, it is the most expensive edge, so
crossing it should *dominate* any path that uses it.

Comparing the two readings' path lengths directly would be meaningless — one is
measured in units of confidence and the other in units of 1/confidence. What is
comparable is the share of a crossing path that the bridge accounts for:

```{r two-readings}
bridge_share <- function(type) {
  w <- as_netkit_weights_demo(g, "confidence", type)
  # Cost of getting from one clique to the other, and the bridge's own cost.
  total <- distances(g, v = "L2", to = "R2", weights = w$distance)[1, 1]
  bridge <- w$distance[as.numeric(E(g)["L1" %--% "R1"])]
  c(path_cost = total, bridge_cost = bridge, bridge_share = bridge / total)
}

rbind(strength = bridge_share("strength"),
      distance = bridge_share("distance"))
```

Read as a strength, the bridge is the cheapest hop on the route and accounts for
about 5% of it. Read as a cost, the very same number makes it the most expensive
hop, accounting for over 80%. There is no way to get this right by inspection —
both sets of numbers look perfectly plausible.

## The default is to ignore weights, loudly

netkit's `weights` argument defaults to `NULL`, which means **ignore edge
weights entirely**, even when the graph carries a `weight` attribute. When it
finds one it is ignoring, it says so:

```{r warn}
E(g)$weight <- E(g)$confidence

res <- tryCatch(
  summarize_graph_metrics(g),
  warning = function(w) conditionMessage(w)
)
res
```

This differs from plain `igraph`, where the attribute is picked up
automatically. The difference is deliberate: picking it up automatically is how
the three-way inconsistency above stays invisible. Naming the attribute, or
having no `weight` attribute at all, is silent.

## What changes when weights are used

Every function that can use weights takes the same two arguments, and derives
both a strength and a cost vector from them so each metric receives the one it
needs.

```{r weighted-metrics}
set.seed(1)
net <- sample_gnp(60, 0.08, directed = FALSE)
V(net)$name <- paste0("g", seq_len(vcount(net)))
E(net)$confidence <- runif(ecount(net), 0.05, 1)

unweighted <- summarize_graph_metrics(net)
weighted <- summarize_graph_metrics(net, weights = "confidence")

compare <- rbind(unweighted = unweighted, weighted = weighted)
compare[, c("Avg_degree", "Avg_strength", "Average_path_length",
            "Clustering_coefficient", "Avg_betweenness")]
```

`Avg_degree` is always the unweighted count and `Avg_strength` the weighted sum,
so the two rows stay comparable. The path-based columns switch to their weighted
definitions, and `Clustering_coefficient` becomes Barrat's weighted
transitivity.

`node_metrics()` makes the same distinction per vertex:

```{r node-metrics}
nm <- node_metrics(net, metrics = c("degree", "strength", "betweenness"),
                   weights = "confidence", normalized = FALSE, plot = FALSE)
head(nm$result)
```

## Diffusion propagates along strengths

Signal propagation is the one place where there is no ambiguity: a stronger edge
carries more signal. This is also where weights used to be dropped silently, so
it is worth seeing the effect directly.

Back to the barbell, with the bridge almost severed:

```{r barrier}
E(g)$w <- 1
g <- set_edge_attr(g, "w", index = E(g)["L1" %--% "R1"], value = 0.001)

far <- paste0("R", 1:4)

open <- network_diffusion(g, "L2", method = "rwr")
blocked <- network_diffusion(g, "L2", method = "rwr", weights = "w")

c(unweighted = mean(open$score[open$node %in% far]),
  weighted   = mean(blocked$score[blocked$node %in% far]))
```

With every edge treated alike, signal from `L2` reaches the far clique freely.
With the bridge at weight 0.001, almost none of it gets through — which is what
a near-severed connection should do.

A strength of exactly zero means "no connection": it becomes an infinite
distance, which the path-based measures treat as unreachable, and blocks
diffusion completely.

## Seeding from a continuous signal

Related but separate: `seed_weights` lets the diffusion start from values rather
than from set membership. If your seeds came from a differential-expression
result, they are not equally important, and their magnitudes carry the evidence.

```{r seed-weights}
seeds <- c("g1", "g2", "g3")

binary <- network_diffusion(net, seeds, method = "rwr")
graded <- network_diffusion(net, seeds, method = "rwr",
                            seed_weights = c(g1 = 3.1, g2 = 0.4, g3 = 1.2))

head(data.frame(node = binary$node,
                binary = binary$score,
                graded = graded$score[match(binary$node, graded$node)]))
```

## What is not a weight

Signed attributes are rejected rather than coerced:

```{r signed, error = TRUE}
E(net)$effect <- sample(c(-1, 1), ecount(net), replace = TRUE)
summarize_graph_metrics(net, weights = "effect")
```

Activation versus inhibition, or a signed correlation, is not an edge weight —
the reciprocal conversion between strength and distance has no meaning for a
negative number. Take the absolute value if magnitude is what matters, and keep
the sign as a separate edge attribute for annotation and visualization.

## Reusing a diffusion kernel

`prepare_diffusion()` builds the kernel once so repeated runs can skip it. A
kernel is only valid for the weights it was built from, and netkit checks rather
than trusting you to remember:

```{r kernel, error = TRUE}
kernel <- prepare_diffusion(net, method = "rwr", weights = "confidence")

# Correct: same weights as the kernel was built with.
head(network_diffusion(net, seeds, method = "rwr", weights = "confidence",
                       precompute = kernel), 3)

# Rejected: the kernel describes a different graph than the one requested.
network_diffusion(net, seeds, method = "rwr", precompute = kernel)
```

## Summary

| If your edge attribute is | pass |
|---|---|
| a confidence score, correlation, count, interaction score | `weight_type = "strength"` (the default) |
| a cost, a dissimilarity, a transit time | `weight_type = "distance"` |
| a sign (activation/inhibition) | not a weight — use `abs()`, or keep it as annotation |
| not meant to be used at all | nothing; `weights = NULL` is the default |

The one rule worth remembering: **netkit never reads a weight you did not ask it
to read**, and it tells you when a graph carries one you ignored.
